### 2019-01-16 submission/manuscript.tex
Pat's edits

\textbf{\huge Assessing the differences in 16S rRNA gene sequence data that are due to choice of DNA polymerase and number of rounds of PCR}

@@

\textbf{Running title:} Quantifying the choice of DNA polymerase

@@

PCR amplification of 16S rRNA genes is a critical, yet under appreciated
step in the generation of sequence data to describe the taxonomic
composition of microbial communities. Numerous factors in the design of
PCR can impact the sequencing error rate, the abundance of chimeric
sequences, and the degree to which the fragments in the product
represent their abundance in the original sample. The choice of PCR
cycling conditions are known to impact the outcome of PCR and downstream
inferences. We compared the performance of high fidelity polymerases and
varying number of rounds of amplification when amplifying a mock
community with known sequences and community structure and human stool
samples. Although it is impossible to derive specific recommendations,
we did observe general trends. Namely, using a polymerase with the
highest possible fidelity and minimizing the number of rounds of PCR
reduced the sequencing error rate, fraction of chimeric sequences, and
bias. Evidence of bias at the sequence level was subtle and could not be
ascribed to the fragments' fraction of bases that were guanines or
cytosines. When analyzing the mock community data, the amount that the
community deviated from the expected composition increased with rounds
of PCR. The effect was inconsistent for human stool samples. Overall the
results underscore the difficulty of comparing sequence data that are
generated by different PCR protocols; however, the results indicate that
the variation in human stool samples is generally larger than that
introduced by the choice of polymerase or number of rounds of PCR.

@@

The growth in interest in characterizing microbial communities from
diverse communities has exploded with the steep decline in sequencing
costs. A significant amount of effort has gone into understanding the
error profiles of DNA sequencers; however, little work has been done to
understand the effects of choices that researchers make to generate the
PCR amplicons that are sequenced. Here we quantified the effects of the
choice of polymerase and number of PCR cycles on the quality of
downstream data. Overall, we found that these choices can have a
profound impact on the way that a microbial community is represented in
the sequence data. Although the effects are relatively small compared to
the variation in human stool samples, care should be taken to use
polymerases with the highest possible fidelity and to minimize the
number of rounds of PCR.

@@

16S rRNA gene sequencing is a powerful and widely used tool for
surveying the structure of microbial communities (1--3). This approach
has exploded in popularity with the advent of high throughput sequencing
where it is possible to characterize numerous samples with thousands of
sequences per sample. Numerous factors can impact how a natural
community is represented by the sequencing data including the method of
acquiring samples (4--8), storage conditions (4--6, 9--12), extraction
methods (13), amplification conditions (8, 14, 15), sequencing method
(15--17), and analytical pipeline (15, 18--20). The increased sampling
depth that is now available relative to previous Sanger-based methods is
expected to compound the impacts of an investigator's choices and the
interpretation of their results.

One step in the generation of 16S rRNA gene sequence data that has been
long known to have a significant impact on the description of microbial
communities is the choice of conditions for PCR amplification (8, 14,
15). Factors such as the choice of primers have an obvious impact on
which populations will be amplified (18, 21). However, a variety of PCR
artifacts can also impact the perception of a community including the
formation of chimeras (14, 22--24), misincorporation of nucleotides (23,
25, 26), preferential amplification of populations leading to bias (24,
27--33), and accumulation of random amplification events leading to PCR
drift (24, 27, 32, 34). Many bioinformatic tools have been developed to
identify chimeras, there are significant sensitivity and specificity
tradeoffs (14, 35). Others have attempted to account for PCR bias using
modeling approaches (29, 36); however, these have been developed for
idealized situations. Laboratory-based solutions to minimize chimera
formation have also been proposed such as minimizing the amount of
template DNA in the PCR, minimizing the number of rounds of PCR,
minimizing the amount of shearing in the template DNA, and using DNA
polymerases that have a proof-reading ability (14, 23). To minimize PCR
drift, some investigators will pool technical replicate PCRs hoping to
average out the drift (34). Other factors that have been shown to impact
the formation of PCR artifacts are outside the control of an
investigator including the fraction of DNA bases that are guanines or
cytosines, the variation in the length of the targeted region across the
community, the sequence of the DNA that flanks the template, and the
genetic diversity of the community (28, 30--33). Early investigations of
the factors that lead to the formation of PCR artifacts focused on
analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to
explore PCR biases and chimera formation. Although these studies were
instrumental in forcing researchers to be cautious about the
interpretation of their results, we have a poor understanding of how
these factors affect the formation of PCR artifacts in more complex
communities.

The influence that the choice of DNA polymerase has on the formation of
PCR artifacts has not been well studied. There has been recent interest
in how the choice of the hypervariable region and data analysis
pipelines impact the sequencing error rate (15, 18--20); however, these
studies use the same DNA polymerase in the PCR step and implicitly
assume that the rate of nucleotide misincorporation from PCR are
significantly smaller than those from the sequencing phase. There has
been more limited interest in the impact that DNA polymerase choice has
on the formation of chimeras (23, 37). A recent study found differences
in the number of OTUs and chimeras between normal and high fidelity DNA
polymerases (37). The authors of this study could reduce the difference
between two polymerases by optimizing the annealing and extension steps
within the PCR protocol (37). Yet this optimization was specific for the
community they were analyzing (i.e.~captive and semi-captive red-shanked
doucs) and assumed that if the two polymerases generate the same
community structure that the community structure was correct. In fact,
the community structure generated by both methods was not free of
artifacts, but has the same artifacts. A challenge in these types of
experiments is having a priori knowledge of the true community
representation. A mock community with known composition allows
researchers to quantify the sequencing error rate, fraction of chimeras,
and bias (19); however, mock communities have a limited phylogenetic
diversity relative to natural communities. Natural communities, in
contrast, have an unknown community composition making absolute
measurements impossible. They can be used to validate results from mock
communities and to understand the relative impacts of artifacts on the
ability to differentiate biological and methodological sources of
variation. Given the large number of DNA polymerases available to
researchers it is unlikely that a specific recommendation is possible.
Rather, the development of general best practices and understanding the
impact of PCR artifacts on an analysis are desired.

This study investigated the impact of choice of high fidelity DNA
polymerase and the number of rounds of amplification on the formation of
PCR artifacts using a mock community and human stool samples. These PCR
artifacts included (i) the effect of the polymerase on the error rate of
the bases represented in the final sequences, (ii) the fraction of
sequences that appeared to be chimeras and the ability to detect those
chimeras, (iii) the bias of preferentially amplifying one fragment over
another in a mixed pool of templates, and (iv) inter-sample variation in
community structure of samples amplified with the same polymerase across
the amplification process. To characterize these factors we sequenced a
mock community of 8 organisms with known sequences and community
structure and human fecal samples with unknown sequences and community
structures. We sequenced the V4 region of the 16S rRNA genes from a mock
community by generating paired 250 nt reads on the Illumina MiSeq
platform. This region and sequencing approach was used because it has
been shown to result in a relatively low sequencing error rate and is a
widely used protocol. To better understand the impact of DNA polymerase
choice on PCR artifacts, we selected five high fidelity DNA polymerases
and amplified the communities using 15, 20, 25, 30, and 35 rounds of
amplification (Table 1). Collectively, our results suggest that the
number of rounds and to a lesser degree the choice of DNA polymerase
used in PCR impact the sequence data, the effects are consistent and are
smaller than the biological differences between individuals.

@@

\textbf{\emph{Sequencing errors vary by the number of cycles and the DNA
polymerase used in PCR.}} The presence of sequence errors can confound
the ability to accurately classify 16S rRNA gene sequences and group
sequences into operational taxonomic units (OTUs). More importantly,
sequencing errors themselves can alter the representation of the
community. Therefore, it is important to minimize the number of
sequencing errors. Using a widely-used approach that generates the
lowest reported error rate, we quantified the error rate by sequencing
the V4 region of the 16S rRNA genes from an 8 member mock community. We
also removed any contigs that were at least three bases more similar to
a chimera of two references than to a single reference sequence (18, 19,
38). Regardless of the polymerase, the error rate increased with the
number of rounds of amplification (Figure 1). Using 30 rounds of PCR is
a common approach across diverse types of samples. Among the data
generated using 30 rounds of PCR the Accuprime polymerase had the
highest error rate (i.e.~0.124\%) followed by the Platinum
(i.e.~0.094\%), Phusion (i.e.~0.064\%), Kappa (i.e.~0.062\%), and Q5
(i.e.~0.060\%) polymerases (Figure 1). When we applied the
pre.clustering denoising step, which merges the counts of reads within 2
nt of a more abundant sequence, the error rates dropped considerably
such that the Platinum polymerase had the highest error rate
(i.e.~0.014\%) followed by the Accuprime (i.e.~0.012\%), Q5
(i.e.~0.0053\%), Phusion (i.e.~0.0049\%), and Kappa (i.e.~0.0049\%)
polymerases (Figure 1). Although specific recommendations are difficult
to make because the impact of the actual community structure and
concentration of the initial DNA template are likely to have an impact
on the results, it is clear that using as few PCR cycles as necessary
and a polymerase with the lowest possible error rate is a good guide to
minimizing the impact of polymerase on the error rate.

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors.
We identified those contigs that were at least three bases more similar
to a chimera of two references than to a single reference sequence. As
expected from previous studies, the number of chimeras increased with
rounds of amplification (Figure 2A). Interestingly, the fraction of
chimeric sequences from the mock community varied by the type of
polymerase used. After 30 rounds of PCR, the Platinum polymerase had the
highest chimera rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%),
Phusion (i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
polymerases. We used the UChime algorithm to detect chimeras in our
pre-clustered mock community data and calculate the algorithm's
sensitivity and specificity (Figure 2A). For all polymerases, the
specificity was above 95.7\% and showed a weak association with the
number of cycles used. There was considerable inter-polymerase and
inter-round of amplification variation in the sensitivity of UChime to
detect the chimeras from the mock community. This suggested that the
residual error rate after pre-clustering the sequence data did not
compromise the sensitivity. The sensitivity of UChime varied between 50
and 87.0\%. The generalizability of these results is limited because we
used a single mock community with limited genetic diversity. Although we
did not know the true chimera rate for our four human stool samples, we
were able to calculate the fraction of sequences that UChime identified
as being chimeric (Figure 2B). These results followed those from the
mock communities; additional rounds of amplification significantly
increased the rate of chimeras and there was variation between the
polymerases that we used. Although it was not possible to identify the
features of a polymerase that resulted in higher rates of chimeras, it
is clear that using the smallest number of PCR cycles possible will
minimize the impacts of chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to sequence 16S rRNA genes there has
been concern that amplifying fragments from a mixed template pool could
lead to a biased representation in the pool of products and would
confound downstream analyses. The mock community was generated by mixing
equal amounts of genomic DNA from each of the 8 bacteria resulting in
uneven representation of the \emph{rrn} operons across the bacteria. The
vendor of the mock community subjects each lot of genomic DNA to shotgun
sequencing to more accurately quantify the actual abundance of each
organism in the community. We compared the expected relative abundance
of the 16S rRNA genes from each bacterium in the mock community to the
data we generated across rounds of amplification and polymerase (Figure
3). Interestingly, for some bacteria, their representation became less
biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
others became more biased (e.g.~\emph{E. faecalis}), and others had
little change (e.g.~\emph{B. subtilis}). The percentage of bases in the
V4 region that were guanines or cytosines was not predictive of the
amount of bias. Across the strains there was no variation in the length
of their V4 regions and they each had the same sequence in the region
that the primers annealed. One of the bacteria represented in the mock
community, \emph{S. enterica}, had 6 identical copies of the V4 region
and 1 copy that differed from those by one nucleotide. The dominant copy
had a thymidine and the rare copy had a guanine. We used the sequence
data to calculate the ratio of the dominant to rare variants from
\emph{S. enterica} (Figure S1). The Accuprime, Phusion, Platinum, and Q5
polymerases converged to a ratio of 5.4; however, the ratio for the
Kappa polymerase varied between 6.1 and 6.8 for 25 to 35 rounds of PCR.
Given the subtle nature of the variation in the relative abundances of
each 16S rRNA gene fragment, it was not possible to create generalizable
rules that would explain the bias and the previous observation regarding
the guanine and cytosine relative abundance was not observed with our
data.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the distance between
samples and across rounds of PCR could be artificially inflated due to
sequencing errors and chimeras, we used analyzed the alpha and beta
diversity of the mock community data at different phases of the sequence
curation pipeline (Figure 4). First, we removed the chimeras from the
mock community data as described above and mapped the individual reads
to the OTUs that the 16S rRNA gene fragments would cluster into if there
were no sequencing errors. This gave us a community distribution that
reflected the distribution following PCR without any artifacts. Alpha
diversity metrics such as richness and Shannon and Inverse Simpson
diversity metrics increased with the number of rounds of PCR for all
polymerases except for the Kappa polymerase, where the richness remained
constant but the diversity decreased (Figure 4A). These data suggest
that PCR had the effect of making the community distribution more even
than it was originally. Next, we used the true sequence diversity, but
removed chimeras based on mapping reads to all possible chimeras between
reference sequences, and clustered the reads to OTUs. The same trends
were observed as with perfect sequence data except that the richness and
diversity metrics trended higher (Figure 4A). Finally, we used the
observed sequence data and used the UCHIME algorithm to identify
chimeras. Again, we observed the same trends as with the other analyses
except the richness and diversity metrics trended higher (Figure 4A).
These comparisons demonstrated that although the bias at the sequence
level was subtle, PCR does have a bias that is exacerbated by errors and
chimeras. When we measured the Bray-Curtis distance between the
communities observed after 25 rounds of amplification and those at 30
and 35 rounds of amplification the distances between 25 and 35 rounds
were higher than between 25 and 30 rounds for each of the polymerases by
an average of 0.022 units (Figure 4B). The Platinum polymerase varied
the most across rounds of amplification (25 vs 30 rounds: 0.12; 25 vs 35
rounds: 0.15). Although the distances between samples were small, the
ordination of these distances showed a clear change in community
structure with increasing rounds of PCR (Figure 4C). This observation
was supported by our statistical analysis, which revealed that the
number of rounds of PCR (R\textsuperscript{2}=0.21, P\textless{}0.001)
was a larger factor than the choice of polymerase
(R\textsuperscript{2}=0.20, P\textless{}0.001). These results
demonstrate that subtle differences in relative abundances can have an
impact on overall community structure. This variation underscores the
importance of only comparing sequence data that have been generated
using the same PCR conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} The biases that we observed at
the population and community levels using mock community data appeared
to be small relative to the expected differences between biological
samples. To study this further, we calculated alpha and beta-diversity
metrics using the human stool samples for each of the polymerases and
rounds of amplification. We calculated the number of observed OTUs,
Shannon diversity, inverse Simpson diversity index for each condition
and donor (Figure 5A). Although there were clear differences between
conditions, the relative ordering of the stool samples did not
meaningfully vary across conditions. When we characterized the variation
between rounds of amplification using human stool samples, the distance
between the 25 and 30 rounds and 25 and 35 rounds varied considerably
between samples and polymerases (Figure 5B). In general the inter-round
variation was lowest for the data generated using the Kappa and
Accuprime polymerases. The data generated using the Platinum polymerase
was consistent across rounds, but it was more biased than these
polymerases. Considering the average distance across the four samples
varied between 0.39 and 0.56, regardless of the polymerases and number
of rounds of amplification, any bias due to amplification is unlikely to
obscure community-wide differences between samples. In support of this
was our principle coordinates analysis of the Bray-Curtis distances,
which revealed distinct clusters by donor (Figure 5C). Within each
cluster there were no obvious patterns related to the polymerase or
number of rounds of PCR. However, our statistical analysis using adonis
revealed statistically significant differences in the community
structures with the subject explaining the most variation
(R\textsuperscript{2}=0.79, P\textless{}0.001), followed by the number
of rounds of PCR (R\textsuperscript{2}=0.044, P\textless{}0.001) and the
choice of polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001).
Together, these results indicate that for a coarse analysis of
communities, the choice of number of rounds of amplification or
polymerase are not important, but that they must be consistent across
samples. It is difficult to develop a specific recommendation based on
the level of bias across rounds of PCR or polymerases; however, the
general suggestion is to use as few rounds of amplification as possible.

\textbf{\emph{There is little evidence of a relationship between
polymerase or number of rounds of amplification on PCR drift.}} There
have been concerns that the same template DNA subjected to the same PCR
conditions could result in different representations of communities
because of random drift. To test this, we determined the average
Bray-Curtis distance between replicate reactions using the same
polymerase and number of rounds of amplification (Figure 6). Using the
mock community data there were no obvious trends. The average
Bray-Curtis distance within a set of conditions varied by 0.063 to 0.11
units. Although we did not obtain replicates of each of the stool
samples, the intra-sample variation for each set of conditions was
consistent and varied between 0.50 and 0.56 units. These data suggest
that amplicon sequencing is robust to random variation in amplification
and that any differences are likely to be smaller than what is
considered biologically relevant.

@@

\leavevmode\hypertarget{ref-Tidyverse2017}{}%
49. \textbf{Wickham H}. 2017. Tidyverse: Easily install and load the
'tidyverse'.

>>>>

\textbf{\huge Quantifying the impact of DNA polymerase and number of rounds of amplification in PCR on 16S rRNA gene sequence data}

@@

\textbf{Running title:} Quantifying the effects of PCR

@@

PCR amplification of 16S rRNA genes is a critical, yet under appreciated
step in the generation of sequence data to describe the taxonomic
composition of microbial communities. Numerous factors in the design of
PCR can impact the sequencing error rate, the abundance of chimeric
sequences, and the degree to which the fragments in the product
represent their abundance in the original sample. The choice of PCR
cycling conditions are known to impact the outcome of PCR and downstream
inferences. We compared the performance of high fidelity polymerases and
varying number of rounds of amplification when amplifying a mock
community with known sequences and community structure and human stool
samples. Although it was impossible to derive specific recommendations,
we did observe general trends. Namely, using a polymerase with the
highest possible fidelity and minimizing the number of rounds of PCR
reduced the sequencing error rate, fraction of chimeric sequences, and
bias. Evidence of bias at the sequence level was subtle and could not be
ascribed to the fragments' fraction of bases that were guanines or
cytosines. When analyzing mock community data, the amount that the
community deviated from the expected composition increased with rounds
of PCR. The effect was inconsistent for human stool samples. Overall the
results underscore the difficulty of comparing sequence data that are
generated by different PCR protocols. However, the results indicate that
the variation in human stool samples is generally larger than that
introduced by the choice of polymerase or number of rounds of PCR.

@@

A steep decline in sequencing costs drove an explosion in studies
characterizing microbial communities from diverse environments. Although
a significant amount of effort has gone into understanding the error
profiles of DNA sequencers, little has been done to understand the
downstream effects of the PCR amplification protocol. We quantified the
effects of the choice of polymerase and number of PCR cycles on the
quality of downstream data. We found that these choices can have a
profound impact on the way that a microbial community is represented in
the sequence data. The effects are relatively small compared to the
variation in human stool samples, however, care should be taken to use
polymerases with the highest possible fidelity and to minimize the
number of rounds of PCR. These results underscore the difficulty of
directly comparing the sequence data generated using different PCR
protocols or interpreting results in an absolute sense.

@@

16S rRNA gene sequencing is a powerful and widely used tool for
surveying the structure of microbial communities (1--3). This approach
has exploded in popularity with the advent of high throughput sequencing
where it is possible to characterize numerous samples with thousands of
sequences per sample. Numerous factors can impact how a natural
community is represented by the sequencing data including the method of
acquiring samples (4--8), storage conditions (4--6, 9--12), extraction
methods (13), amplification conditions (8, 14, 15), sequencing method
(15--17), and analytical pipeline (15, 18--20). The increased sampling
depth that is now available relative to previous Sanger sequencing-based
methods is expected to compound the impacts of an investigator's choices
and the interpretation of their results.

One step in the generation of 16S rRNA gene sequence data that has been
long known to have a significant impact on the description of microbial
communities is the choice of conditions for PCR amplification (8, 14,
15). Factors such as the choice of primers have an obvious impact on
which populations will be amplified (18, 21). However, a variety of PCR
artifacts can also impact the perception of a community including the
formation of chimeras (14, 22--24), misincorporation of nucleotides (23,
25, 26), preferential amplification of populations leading to bias (24,
27--33), and accumulation of random amplification events leading to PCR
drift (24, 27, 32, 34). Many bioinformatic tools have been developed to
identify chimeras, there are significant sensitivity and specificity
tradeoffs (14, 35). Laboratory-based solutions to minimize chimera
formation have also been proposed such as minimizing the amount of
template DNA in the PCR, minimizing the number of rounds of PCR,
minimizing the amount of shearing in the template DNA, and using DNA
polymerases that have a proof-reading ability (14, 23). Others have
attempted to account for PCR bias using modeling approaches (29, 36);
however, these have been developed for idealized situations. To minimize
PCR drift, some investigators will pool technical replicate PCRs hoping
to average out the drift (34). Other factors that have been shown to
impact the formation of PCR artifacts are outside the control of an
investigator including the fraction of DNA bases that are guanines or
cytosines, the variation in the length of the targeted region across the
community, the sequence of the DNA that flanks the template, and the
genetic diversity of the community (28, 30--33). Early investigations of
the factors that lead to the formation of PCR artifacts focused on
analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to
explore PCR biases and chimera formation. Although these studies were
instrumental in forcing researchers to be cautious about the
interpretation of their results, we have a poor understanding of how
these factors affect the formation of PCR artifacts in more complex
communities.

The influence that the choice of DNA polymerase has on the formation of
PCR artifacts has not been well studied. There has been recent interest
in how the choice of the hypervariable region and data analysis
pipelines impact the sequencing error rate (15, 18--20); however, these
studies use the same DNA polymerase in the PCR step and implicitly
assume that the rate of nucleotide misincorporation from PCR are
significantly smaller than those from the sequencing phase. There has
been more limited interest in the impact that DNA polymerase choice has
on the formation of chimeras (23, 37). A recent study found differences
in the number of OTUs and chimeras between normal and high fidelity DNA
polymerases (37). The authors of the study reduced the difference
between two polymerases by optimizing the annealing and extension steps
within the PCR protocol (37). Yet this optimization was specific for the
community they were analyzing (i.e.~captive and semi-captive red-shanked
doucs) and assumed that if the two polymerases generate the same
community structure that the community structure was correct. In fact,
the community structure generated by both methods was not free of
artifacts, but likely had the same artifacts. A challenge in these types
of experiments is having \emph{a priori} knowledge of the true community
representation. A mock community with known composition allows
researchers to quantify the sequencing error rate, fraction of chimeras,
and bias (19); however, mock communities have a limited phylogenetic
diversity relative to natural communities. Natural communities, in
contrast, have an unknown community composition making absolute
measurements impossible. They can be used to validate results from mock
communities and to understand the relative impacts of artifacts on the
ability to differentiate biological and methodological sources of
variation. Given the large number of DNA polymerases available to
researchers, it is unlikely that a specific recommendation is possible.
Rather, the development of general best practices and understanding the
impact of PCR artifacts on an analysis are needed.

This study investigated the impact of choice of high fidelity DNA
polymerase and the number of rounds of amplification on the formation of
PCR artifacts using a mock community and human stool samples. These PCR
artifacts included (i) the effect of the polymerase on the error rate of
the bases represented in the final sequences, (ii) the fraction of
sequences that appeared to be chimeras and the ability to detect those
chimeras, (iii) the bias of preferentially amplifying one fragment over
another in a mixed pool of templates, and (iv) inter-sample variation in
community structure of samples amplified with the same polymerase across
the amplification process. To characterize these factors we sequenced a
mock community of 8 organisms with known sequences and community
structure and human fecal samples with unknown sequences and community
structures. We sequenced the V4 region of the 16S rRNA genes from a mock
community by generating paired 250 nt reads on the Illumina MiSeq
platform. This region and sequencing approach was used because it has
been shown to result in a relatively low sequencing error rate and is a
widely used protocol ({\textbf{???}}). To better understand the impact
of DNA polymerase choice on PCR artifacts, we selected five high
fidelity DNA polymerases and amplified the communities using 20, 25, 30,
and 35 rounds of amplification. Collectively, our results suggest that
the number of rounds and to a lesser degree the choice of DNA polymerase
used in PCR impact the sequence data, the effects are consistent and are
smaller than the biological differences between individuals.

@@

\textbf{\emph{Sequencing errors vary by the number of cycles and the DNA
polymerase used in PCR.}} The presence of sequence errors can confound
the ability to accurately classify 16S rRNA gene sequences and group
sequences into operational taxonomic units (OTUs). More importantly,
sequencing errors themselves can alter the representation of the
community. Therefore, it is important to minimize the number of
sequencing errors. Using a widely-used approach that generates the
lowest reported error rate, we quantified the error rate by sequencing
the V4 region of the 16S rRNA genes from an 8 member mock community. We
also removed any contigs that were at least three bases more similar to
a chimera of two references than to a single reference sequence (18, 19,
38). Regardless of the polymerase, the error rate increased with the
number of rounds of amplification (Figure 1). Using 30 rounds of PCR is
a common approach across diverse types of samples. Among the data
generated using 30 rounds of PCR the Accuprime polymerase had the
highest error rate (i.e.~0.124\%) followed by the Platinum
(i.e.~0.094\%), Phusion (i.e.~0.064\%), Kappa (i.e.~0.062\%), and Q5
(i.e.~0.060\%) polymerases (Figure 1). When we applied a pre-clustering
denoising step, which merged the counts of reads within 2 nt of a more
abundant sequence (19), the error rates dropped considerably such that
the Platinum polymerase had the highest error rate (i.e.~0.014\%)
followed by the Accuprime (i.e.~0.012\%), Q5 (i.e.~0.0053\%), Phusion
(i.e.~0.0049\%), and Kappa (i.e.~0.0049\%) polymerases (Figure 1).
Although specific recommendations are difficult to make because the
phylogenetic diversity of the initial DNA template is likely to have an
impact on the results, it is clear that using as few PCR cycles as
necessary and a polymerase with the lowest possible error rate is a good
guide to minimizing the impact of polymerase on the error rate.

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors
(14, 35). Because the true sequences of the organisms in the mock
community were known, we could generate all possible chimeras between
pairs of V4 16S rRNA gene fragments and use these possible chimeric
sequences to screen the sequences generated under the different PCR
conditions for chimeras. The number of chimeras increased with rounds of
amplification (Figure 2A). Interestingly, the fraction of chimeric
sequences from the mock community varied by the type of polymerase used.
After 30 rounds of PCR, the Platinum polymerase had the highest chimera
rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%), Phusion
(i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
polymerases. Because our chimera screening procedure can only be applied
to mock communities, we used the UCHIME algorithm to model the chimera
screening approach that is used in most sequence curation pipelines. By
comparing the output of UCHIME to our reference-based approach, we were
able to calculate the UCHIME's sensitivity and specificity (Figure 2A).
The specificity for all polymerases was above 95.7\% and showed a weak
association with the number of cycles used. There was considerable
inter-polymerase and inter-round of amplification variation in the
sensitivity of UCHIME to detect the chimeras from the mock community.
This suggested that the residual error rate after pre-clustering the
sequence data did not compromise the sensitivity. The sensitivity of
UCHIME varied between 50 and 87.0\%. The generalizability of these
results is limited because we used a single mock community with limited
genetic diversity. Although we did not know the true chimera rate for
our four human stool samples, we were able to calculate the fraction of
sequences that UCHIME identified as being chimeric (Figure 2B). These
results followed those from the mock communities: additional rounds of
amplification significantly increased the rate of chimeras and there was
variation between the polymerases that we used. Although it was not
possible to identify the features of a polymerase that resulted in
higher rates of chimeras, it is clear that using the smallest number of
PCR cycles possible will minimize the impacts of chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to amplify 16S rRNA gene fragments
there has been concern that amplifying fragments from a mixed template
pool could lead to a biased representation in the pool of products and
would confound downstream analyses (24, 27--33). The mock community was
generated by mixing equal amounts of genomic DNA from each of the 8
bacteria resulting in uneven representation of the \emph{rrn} operons
across the bacteria because each bacterium had a different genome size
and varied in the number of operons in its genome. The vendor of the
mock community subjects each lot of genomic DNA to shotgun sequencing to
more accurately quantify the actual abundance of each organism in the
community. We compared the expected relative abundance of the 16S rRNA
genes from each bacterium in the mock community to the data we generated
across rounds of amplification and polymerase (Figure 3). Interestingly,
for some bacteria, their representation became less biased with
additional rounds of PCR (e.g.~\emph{L. fermentum}), while others became
more biased (e.g.~\emph{E. faecalis}), and others had little change
(e.g.~\emph{B. subtilis}). Contrary to prior reports (28), The
percentage of bases in the V4 region that were guanines or cytosines was
not predictive of the amount of bias. Across the strains there was no
variation in the length of their V4 regions and they each had the same
sequence in the region that the primers annealed. One of the bacteria
represented in the mock community, \emph{S. enterica}, had 6 identical
copies of the V4 region and 1 copy that differed from those by one
nucleotide. The dominant copy had a thymidine and the rare copy had a
guanine. We used the sequence data to calculate the ratio of the
dominant to rare variants from \emph{S. enterica} (Figure S1). The
Accuprime, Phusion, Platinum, and Q5 polymerases converged to a ratio of
5.4; however, the ratio for the Kappa polymerase varied between 6.1 and
6.8 for 25 to 35 rounds of PCR. Given the subtle nature of the variation
in the relative abundances of each 16S rRNA gene fragment, it was not
possible to create generalizable rules that would explain the bias.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the variation in bias
between polymerases and across rounds of PCR could be artificially
inflated due to sequencing errors and chimeras, we analyzed the alpha
and beta diversity of the mock community data at different phases of the
sequence curation pipeline (Figure 4). First, we removed the chimeras
from the mock community data as described above and mapped the
individual reads to the OTUs that the 16S rRNA gene fragments would
cluster into if there were no sequencing errors. This gave us a
community distribution that reflected the distribution following PCR
without any artifacts. Although the richness did not change, the Shannon
diversity increased with the number of rounds of PCR for all polymerases
except for the Kappa polymerase, for which the diversity decreased
(Figure 4A). These data suggest that PCR had the effect of making the
community distribution more even than it was originally, except for the
data generated using the Kappa polymerase where the evenness decreased.
Next, we used the observed sequence errors, but removed chimeras using
our reference-based approach, and clustered the reads to OTUs. The
richness and diversity metrics trended higher with higher error rates
and number of rounds of PCR (Figure 4A). Finally, we used the observed
sequence data and used the UCHIME algorithm to identify chimeras. Again,
the richness and diversity metrics trended higher with higher error
rates and number of rounds of PCR (Figure 4A). These comparisons
demonstrated that although the bias at the sequence level was subtle,
PCR introduces bias that is exacerbated by errors and chimeras when
sequences are clustered into OTUs. When we measured the Bray-Curtis
distance between the communities observed after 25 rounds of
amplification and those at 30 and 35 rounds of amplification the
distances between 25 and 35 rounds were higher than between 25 and 30
rounds for each of the polymerases by an average of 0.022 units (Figure
4B). The Platinum polymerase varied the most across rounds of
amplification (25 vs 30 rounds: 0.12; 25 vs 35 rounds: 0.15). Although
the distances between samples were small, the ordination of these
distances showed a clear change in community structure with increasing
rounds of PCR (Figure 4C). This observation was supported by our
statistical analysis, which revealed that the number of rounds of PCR
(R\textsuperscript{2}=0.21, P\textless{}0.001) was a larger factor than
the choice of polymerase (R\textsuperscript{2}=0.20, P\textless{}0.001).
These results demonstrate that subtle differences in relative abundances
can have an impact on overall community structure. This variation
underscores the importance of only comparing sequence data that have
been generated using the same PCR conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} The biases that we observed at
the population and community levels using mock community data appeared
to be small relative to the expected differences between biological
samples. To study this further, we calculated alpha and beta-diversity
metrics using the human stool samples for each of the polymerases and
rounds of amplification. We calculated the number of observed OTUs and
Shannon diversity for each condition and subject (Figure 5A). Although
there were clear differences between conditions, the relative ordering
of the stool samples did not meaningfully vary across conditions. When
we characterized the variation between rounds of amplification using
human stool samples, the distance between the 25 and 30 rounds and 25
and 35 rounds varied considerably between samples and polymerases
(Figure 5B). In general the inter-round variation was lowest for the
data generated using the Kappa and Accuprime polymerases. The data
generated using the Platinum polymerase was consistent across rounds,
but overall, it was more biased than the other polymerases. Considering
the average distance across the four samples varied between 0.39 and
0.56, regardless of the polymerases and number of rounds of
amplification, any bias due to amplification is unlikely to obscure
community-wide differences between samples. In support of this was our
principle coordinates analysis of the Bray-Curtis distances, which
revealed distinct clusters by donor (Figure 5C). Within each cluster
there were no obvious patterns related to the polymerase or number of
rounds of PCR. However, our statistical analysis revealed statistically
significant differences in the community structures with the subject
explaining the most variation (R\textsuperscript{2}=0.79,
P\textless{}0.001), followed by the number of rounds of PCR
(R\textsuperscript{2}=0.044, P\textless{}0.001) and the choice of
polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001). Together,
these results indicate that for a coarse analysis of communities, the
choice of number of rounds of amplification or polymerase are not
important, but that they must be consistent across samples. It is
difficult to develop a specific recommendation based on the level of
bias across rounds of PCR or polymerases; however, the general
suggestion is to use as few rounds of amplification as possible.

\textbf{\emph{There is little evidence of a relationship between
polymerase or number of rounds of amplification on PCR drift.}} There
have been concerns that the same template DNA subjected to the same PCR
conditions could result in different representations of communities
because of random drift over the course of PCR. To test this, we
determined the average Bray-Curtis distance between replicate reactions
using the same polymerase and number of rounds of amplification (Figure
6). Using the mock community data there were no obvious trends. The
average Bray-Curtis distance within a set of conditions varied by 0.063
to 0.11 units. Although we did not generate technical replicates of each
of the stool samples, the inter-sample variation for each set of
conditions was consistent and varied between 0.50 and 0.56 units. These
data suggest that amplicon sequencing is robust to random variation in
amplification and that any differences are likely to be smaller than
what is considered biologically relevant.

@@
